A pipeline to analyse time-course gene expression data

### 2020-06-06 scripts/manuscript.Rmd
updates to text, figure legends

In the following plot, we plot the two first PC components in the first column, where each sample is colored by its
condition (top row) and its sampling time (bottom row). 
We use different plotting symbols for each
replicate. We also plot the second and third components in the second column.

@@

Then, we plot the pearson correlation between each sample as a heatmap diagram (using the function `aheatmap` in `NMF`). We order the
samples by their Group (treatment) and Timepoint (the time of sampling).
Additionally, in this example, we order each treatment by strength of the
pathogenicity of the treatment: Control, Kawasaki, California, low dose of
Vietnam, then high dose of Vietnam.

@@

1) Per-time point analysis, where we consider each time point a different
condition and determine what genes are changing between time points, or
between conditions at a single time-point.

@@

The per-time point analysis is using classical differential expression
approaches, and is often the approach advocated when dealing with small
time-course datasets, where there are only a few time points [@ritchie:limma;
@robinson:edgeR; @love:moderated] . For long time-course datasets, however, a
separate test for each time-point results in creating many different tests,
for example one for every time point, the results of which are difficult to
integrate. We find in practice that the global analysis simplifies analysis
and interpretation of longer time courses data, with per-time point analysis
reserved for particularly interesting comparisons of individual time-points.

@@

**Use with `moanin`** `moanin` provides functionality for performing both of
these types of approaches, though our focus is on the global approach. In both
situations, we first need to set up a object (a `moanin_model` object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, …)

We start by creating the `moanin_model` object using the `create_moanin_model`
function. We need to provide two things to the function: a `data.frame` with
the metadata, and the number of degrees of freedom of the splines used in the
functional modeling. The metadata `data.frame` object should contain at least
two columns: one named `Group`, containing the  treatement effect, and a
second one named `Timepoint` containing the timepoint information.

@@

The `moanin_model` object contains a number of metadata and options used
throughout this analysis: the condition and timepoints of each samples, the
formula object used, the basis matrix, and the degrees of freedom of the model.

If no formula is provided, it will default to the following, which provides
for a different spline fit for every group (as defined by the `Group` variable
in the `meta` data.frame): 
`formula = ~Group:ns(Timepoint, df=degrees_of_fredoom) + Group + 0`.

The basis matrix contains the value of all of the spline basis function for
each sample of the moment. As such, replicate samples will have the same
values.

@@

**EAP: General comment: you need a few more print() and head() of some objects
to demonstrate what the output looks like**

@@

`moanin` provides a simple interface to perform a timepoint by timepoint
differential expression analysis. Comparison between groups is traditionally
done by defining the group comparisons (called `contrasts` in linear models)
as a linear combination of the coefficients of the model.  Comparing groups
within each timepoint can create many contrasts, and thus `moanin` provides
functionality to create these contrasts in an automatic way, and then calls
`limma` [@ritchie:limma] on the set of contrasts provided. By default,
`moanin` expects RNA-Seq contact counts, and will estimate voom weights.

Here, we show an example where we define our contrasts to be the difference
between the control mouse ("M") and the mouse infected with the high dose of
the influenza strain A/Vietnam/1203/04 (H5N1) ("VL") for each time point, but
the function works with any form contrasts [@ritchie:limma].

@@

This creates a vector of contrasts to be tested, one for each timepoint, in the format required by `limma`:

```{r echo=FALSE}
kable(contrasts, "latex", col.names="Contrasts", booktabs=TRUE)
```

@@

```{r echo=FALSE}
kable(weekly_de_analysis[1:10, 1:6 ], "latex", digits=3, booktabs=TRUE)

@@

Let's look at the distribution of genes found differentially expressed per
week between control and each of the influenza strains.

```{r fig.width=6, fig.height=6, echo=FALSE}
timely_barplot = function(de_analysis, 
    labels=NULL, qval_threshold=0.05,
    xlab="Timepoint", ylab="Number of DE genes", main="", ylim=c(0, 20000), col="black", ...){

@@

Such an analysis can demonstrate some general trends, with clearly more genes
being differentially expressed at later time points, and the Vietnam high-dose
showing perhaps an earlier onset than the low-dose.

@@

Another difficulty with such an approach is making sense of the general
temporal structure for any particular gene, as different genes will have
different combinations of timepoints DE. For the comparison of the Kawasaki
strain to the control, for example, there are `r length(signifCombos)` genes
found DE in some timepoints, and there are `r length(tabCombos)` different
combinations of timepoints for which they are DE. Some of these make sense,
such as DE in timepoints 48H-168H (`r tabCombos["48,60,72,120,168"]` genes), but
many are very fragmentary. For example there are `r tabCombos["48,60,120,168"]`
genes which are DE in timepoints 48,60,120,168H, but *not* in the 72H. Many of
these genes are likely to have not made the cutoff for significance in 72H,
but don't show real differences in the overall trend between 48H-168H.

Here are plots of the first 10 such genes, where we can see that some might
show some meaningful changes between 60 and 72H, but others clearly just have
a single replicate that is different or increased variability at 72H.

```{r showMissingDEGenes, echo=FALSE}
exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
plot_splines_data(data, moanin_model, subset_data=exampleGenes,
		  colors=ann_colors$Group, smooth=TRUE)
```

@@

`moanin` extends this idea by providing functionality to compare time course
data between different treatment conditions, using a similar mechanism of
contrasts -- only now the contrasts are differences between the estimated mean
functions. This is done with the function `DE_timecourse`, which takes as
similar input that of `DE_timepoints`, only now will test the entire mean
function (and therefore does not require a step of expanding the contrasts
into contrasts for individual timepoints).

@@

# The function takes the data (data.frame or named matrix), the meta data
# (data.frame containing a timepoint and group column, the first corresponding
# to the time-course information, the latter corresponding to the
# treatment).
DE_results = DE_timecourse(
    data, moanin_model, timecourse_contrasts,
    use_voom_weights=FALSE)
pval_columns = colnames(DE_results)[
    grepl("pval", colnames(DE_results))]
qval_columns = colnames(DE_results)[
    grepl("qval", colnames(DE_results))]
pvalues = DE_results[, pval_columns]
qvalues = DE_results[, qval_columns]
```

@@

```{r echo=FALSE, fig.width=3, fig.height=3, fig.cap="Distribution of DE genes per condition"}
num_de = colSums(qvalues < 0.05)
names(num_de) = c(
    "Kawasaki", "California", "Vietnam LD", "Vietname HD")
barplot(num_de, col="black", ylab="Number of DE genes")
```

The next step in a classical differential expression analysis is typically to
assess the effect of the treatment by looking at the log fold change.
Computing the log fold change on a time-course experiment is not trivial: one
can be interested in the average log-fold change across time, or the
cumulative log-fold change. Sometimes a gene can be over-expressed at the
beginning of the time-course data, and then over-expressed at the end of the
experiment. As a result, `moanin` provides a number of possible ways to
compute the log fold change across the whole time-course.

First, `moanin` provides as simple interface to compute the log-fold change
for each individual timepoints.

@@

Sometimes, a single value per gene and per contrast is more useful. Here is a
table of the possible ways to compute log-fold change values.

| Name   | Formula   |     |
|------- | --------- |--- |
| timely | $\text{lfc}(t)$  | Function of time |
| sum  | $\sum_t \text{lfc}(t)$  | Sum of log fold change. |
| abs_sum  | $\sum_t \lvert \text{lfc}(t)\lvert$  | Always positive |
| max  | $\max_t \lvert \text{lfc}(t) \lvert$  | Always positive  |
| min  | $\min_t \lvert \text{lfc}(t) \lvert$  | Always positive |
| timecourse | See details below | Captures overall strength and direction  |

To capture the overall strength and direction of the response, we leverage the
timepoint by timepoint log-fold change $\text{lfc}(t)$, and apply the
following formula:

@@

Note that when a gene is not consistently up or down-regulated the estimation
of the direction will not represent accurately the changes observed.

To compute the log-fold change, `moanin` provides a function taking as
argument the data, the `moanin_model` object, the contrasts to evaluate, as
well as the method to use to estimate the log-fold change.

@@

```{r display_lfc}
kable(log_fold_change_timecourse[1:5, ], digits=3, booktabs=TRUE)
```

Each methods of computing the log-fold change captures different elements of
the time-course data: overall change, largest change, ...

```{r echo=FALSE, fig.height=6, fig.width=6, message=FALSE}
par(mfrow=c(2, 2))
plot(log_fold_change_timecourse[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="timecourse vs sum",
     xlab="timecourse", ylab="sum")
plot(log_fold_change_timecourse[, "M-C"],
     log_fold_change_max[, "M-C"], pch=20, main="timecourse vs max",
     xlab="timecourse", ylab="max")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_min[, "M-C"], pch=20, main="max vs min",
     xlab="max", ylab="min")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="max vs sum",
     xlab="max", ylab="sum")
```

From the single measures of log-fold change and the p-value, we can now look
at the volcano plot. Here is an example of a volcano plot for the comparaison
of the control to the Kawasaki strain, using the timecourse log fold change
computation.

```{r fig.height=4, fig.width=4}
pvalue = DE_results[, "M-K_pval"]
names(pvalue) = row.names(DE_results)
lfc_timecourse = log_fold_change_timecourse[, "M-K"]
names(lfc_timecourse) = row.names(log_fold_change_timecourse)

@@

As another sanity check, `moanin` provides a simple utility function to
visualize gene time-course data in different conditions. Here, we plot the 10
genes with the smallest p-values.

```{r fig.height=6, fig.width=6}
top_DE_genes_pval = names(sort(pvalue)[1:10])
plot_splines_data(data[top_DE_genes_pval, ], moanin_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))
```

And here we visualize the genes with the largest absolute timecourse log-fold
change.

```{r fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_timecourse),
	 decreasing=TRUE)[1:10])
plot_splines_data(data[top_DE_genes_lfc, ], moanin_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))

@@

Thanks to those visualization, we can see that genes often follow similar
patterns of expression, although on a different scale for each gene. We can
leverage this observation to cluster the genes into groups of similar patterns
of transcriptomic response.

@@

The very large number of genes found differentially expressed impair any
interpretation one would attempt: with 70\% of the genome found differentially
expressed, all pathways are affected by the treatment. Hence the next step of
the workflow to cluster gene expression according to their dynamical response
to the treatment. Before clustering the genes, we first reduce the set of
genes of interest to genes that are (1) found significantly differentially
expressed; (2) a large-fold change between conditions. To do this, we first
aggregate all p-values obtained during the time-course differential expresison
step in a single p-value using Fisher's method [@fisher:statistical]. Then we
select all the genes which have a Fisher adjusted p-value below 0.05 and a log
fold change of at least two between at least one condition and one time-point.
Reducing the set of genes on which to perform the clustering allows to
estimate the centroids with more stability.

@@

After filtering, we are left with `r dim(y)[1]` genes. We can then apply a
clustering. As observed by looking at genes found differentially expressed, 
many genes share a similar gene expression pattern, but on different scale.
We thus propose the following adaptation of k-means:

- **Splines estimation**: for each gene, fit the splines function with the basis
  of your choice (provided in the `moanin_model` object).
- **Rescaling splines**: for each gene, rescale the estimated splines function
  such that the values are bounded between 0 and 1 and thus comparable between genes.
- **K-means**: apply k-means on the rescaled fitted values of the splines to estimate the
  cluster centroids.
- **Assign scores and labels to all genes**: Our
  `splines_kmeans_score_and_label` function assigns a score to all
  gene-cluster pairs based on a goodness-of-fit measure of the gene to the
  cluster centroid. We then "assign" each gene to a cluster based on its
  score.

The first three steps are performed internally by the `splines_kmeans`
function. For now, we will set the number of clusters to be $8$, though we
will return to the question of picking the best number of clusters below.

@@

- `centroids`: a matrix containing the cluster centroids. The matrix is of
  shape (`n_centroid`, `n_samples`).
- `clusters`: a vector of size `n_genes`, containing the

We then use the `plot_splines_data` function to visualize the centroids of
each cluster obtained with the splines k-means model.

@@

These centroids are on a 0-1 scale, because of our rescaling of the spline
fits, and do not represent the actual gene expression level. We can plot a few
of the actual genes in a cluster to see the difference:

```{r fig.height=6, fig.width=6}
cluster_of_interest = 2
cluster2Genes = names(
    kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
dataToPlot=rbind(y[cluster2Genes[3:6], ],"Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
plot_splines_data(dataToPlot, moanin_model,
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

The spline plots are fitted per gene in the above plot, and illustrate that
the genes assigned to a cluster have differing degrees to which their data
fits that of the cluster centroid, and indeed fits that of their individual
spline fits.

@@

Since we arbitrarily chose a subset of genes based on our filter, we would
like all genes to have a chance of assignment to a cluster, as well as a score
for whether it is a good assignment.  The scoring and label step allows us to
assign genes to clusters that were removed during the filtering step, yet are
good matches to the centroids found during the clustering. The score
corresponds to a goodness-of-fit between each gene and each cluster, computed
as follows:

@@

The scoring and labeling is done via the `splines_kmeans_score_and_label`
function. This function calculates the goodness of fit of the gene to the
cluster centroid. By default, the function labels only the 50% of genes with
the best scores, with the remaining genes getting `NA` as their assignment.

@@

    - `scores_and_labels$scores`: the matrix of shape `n_cluster` x `n_genes`,
      containing for each gene and each cluster, the goodness of fit as
      described above.
    - `scores_and_labels$labels`: the labels for all of the genes with a
      good-enough goodness-of-fit score.

First, let us visualize the distribution of goodness-of-fit scores for each
cluster. Here, we display the whole set of scores assigned to each cluster,
prior to filtering poor-quality fits.

```{r visualize_scores, fig.height=6, fix.width=6, echo=TRUE, caption="Distribution of goodness-of-fit scores for each cluster. The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
# Get the best score and best label for all of the genes
scores = rowMin(scores_and_labels$scores)
labels = apply(scores_and_labels$scores, 1, which.min)
max_score = quantile(scores, 0.5)
par(mfrow=c(3, 3))
n_clusters = dim(kmeans_clusters$centroids)[1]
for(cluster_id in 1:n_clusters){
    hist(scores[labels == cluster_id],
	 breaks=(1:50/50), xlim=c(0, 1),
	 col="black", main=paste("C", cluster_id, sep=""),
	 xlab="score", ylab="Num. genes")
    abline(v=max_score, col="red", lwd=3, lty=2)
}

Note that for some of the genes, the best scores is 1. A score of 1 indicates
that there is no correlation between the gene and the cluster: those genes fit
poorly to all clusters and the assignment to one cluster is done randomly..
This underlines the importance of filtering genes that fit poorly to clusters.

@@

After running `scores_and_labels` we now have `r length(labels)` genes that
are assigned to a cluster.

Before performing the next steps, let us investigate in more detail the
differences between the labels provided by the splines k-means model and the
scoring and labeling step.

```{r echo=FALSE, fig.height=6, fig.width=6, warning=FALSE}
par(mfrow=c(2, 1))
barplot(table(kmeans_clusters$clusters),
	col="black",
	main="labeling from splines k-means")
barplot(table(labels), col="black", main="labeling from scores_and_labels function")
```

Of the original `r length(kmeans_clusters$clusters)` genes used in the
clustering, `r length(intersect(names(kmeans_clusters$clusters),
names(labels)))` are still assigned to a cluster based on their goodness of
fit score. We can compare whether they are still assigned to the same clusters
based on a confusion matrix shown below. The kmeans cluster assignments are
shown on the rows, and the goodness-of-fit assignments in the columns, with
the number in each cell indicating the number of genes in the intersection of
the two clusters.

```{r echo=FALSE, fig.height=6, fig.width=6, fig.cap="Confusion matrix between the two sets of labels"}
genes_scored = names(labels)

>>>>

In Figure \ref{fig:pcaPlots}, we visualize the data by plotting the PC components, with samples  colored by either its
condition (top row) or its sampling time (bottom row) and each replicate a different symbol. We can see a large difference in later time points.

@@

We also plot in Figure \ref{fig:correlationPlot} the pearson correlation between each sample as a heatmap diagram (using the function `aheatmap` in `NMF`). We order the
samples by their Group (treatment) and Timepoint (the time of sampling).

@@

1) Per-time point analysis, where we consider each time point a different
condition and determine what genes are changing between specific time points, or
between conditions at a single time-point.

@@

The per-time point analysis is using classical differential expression
approaches, and is often the approach advocated when dealing with small
time-course datasets, where there are only a few time points [@ritchie:limma;
@robinson:edgeR; @love:moderated]. For long time-course datasets, however, a
separate test for each time-point results in creating many different tests,
for example one for every time point, the results of which are difficult to
integrate. We find in practice that the global analysis simplifies analysis
and interpretation of longer time courses data, with per-time point analysis
reserved for particularly interesting comparisons of individual time-points.

@@

**DE Analysis in `moanin`** Our package `moanin` provides functionality for performing both of
these types of approaches, though our focus is on the global approach, specifically by fitting spline models to the genes.

In both
situations, we first need to set up a object (a `moanin_model` object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, ...). The `moanin_model` object
will contain information used
throughout this analysis, in particular the condition and timepoints of each sample and the
the basis matrix for fitting splines models.

We start by creating the `moanin_model` object using the `create_moanin_model`
function. We need to provide two things to the function: a `data.frame` with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata `data.frame` object should contain at least
two columns: one named `Group`, containing the  treatement effect, and a
second one named `Timepoint` containing the timepoint information.

@@

The main operation of the `create_moanin_model` function, in addition to holding the necessary meta data of the samples,
is to create a basis matrix for the splines fit. This matrix gives the evaluation of all of the spline basis functions for
each sample (as such, replicate samples will have the same
values). By default, `create_moanin_model` will create the spline basis functions which will lead to a different spline fit for every group (as defined by the `Group` variable
in the `meta` data.frame). This is done by the following `R` formula syntax:

`formula = ~Group:ns(Timepoint, df=degrees_of_fredoom) + Group + 0`.

Alternatively, the user can provide a formula of their own, or simply provide the basis matrix themselves.

@@

**EAP: This needs to be worked on, because too big to print. Really, need a `print.moanin_model` in the package (submitted issue) and put that here.**

@@

`moanin` provides a simple interface to perform a timepoint by timepoint
differential expression analysis. Comparison between groups is traditionally
done by defining the group comparisons (called `contrasts` in linear models)
as a linear combination of the coefficients of the model.  Comparing groups
within each timepoint can create many contrasts, and thus `moanin` provides
functionality to create these contrasts in an automatic way, and then calls
`limma` [@ritchie:limma] on the set of contrasts provided. By default,
`moanin` expects RNA-Seq gene counts, and will estimate voom weights.

Here, we show an example where we create contrasts that will be be the difference
between the control mouse ("M") and the mouse infected with the high dose of
the influenza strain A/Vietnam/1203/04 (H5N1) ("VL") for each time point (but
the function works with any form contrasts [@ritchie:limma]).

@@

This creates a character vector of contrasts to be tested, one for each timepoint, in the format required by `limma`:

```{r echo=FALSE}
contrasts[1:5]
```
**EAP**: I think it's better here to actually show the output, rather than a table, so you can see what the contrasts actually look like. I like to reserve table format for actual results, rather than code-like input

@@

```{r echo=FALSE}
kable(weekly_de_analysis[1:10, 1:3 ], "latex", digits=3, booktabs=TRUE)

@@

In Figure \ref{fig:timeSignifPerTime}, we show the distribution of genes found differentially expressed per
week between control and each of the influenza strains. Such an analysis can demonstrate some general trends, with clearly more genes
being differentially expressed at later time points, and the Vietnam high-dose
showing perhaps an earlier onset than the low-dose.

```{r timeSignifPerTime,fig.width=6, fig.height=6, echo=FALSE, fig.cap="Number of genes found DE in each timepoint when comparing the influenza strains to control."}
timely_barplot = function(de_analysis, 
    labels=NULL, qval_threshold=0.05,
    xlab="Timepoint", ylab="Number of DE genes", main="", ylim=c(0, 20000), col="black", ...){

@@

@@

Another difficulty with such an approach is making sense of the general
temporal structure for any particular gene, as different genes will have
different combinations of timepoints DE. For the comparison of the Kawasaki
strain to the control, for example, there are `r length(signifCombos)` genes
found DE in some timepoints, and there are `r length(tabCombos)` different
combinations of timepoints for which they are DE. Some of these make sense,
such as DE in timepoints 48H-168H (`r tabCombos["48,60,72,120,168"]` genes), but
many are very fragmentary. For example there are `r tabCombos["48,60,120,168"]`
genes which are DE in timepoints 48,60,120,168H, but *not* in the 72H. Many of
these genes are likely to have not made the cutoff for significance in 72H,
but don't show real differences in the overall trend between 48H-168H. In Figure \ref{fig:showMissingDEGenes} we show the plots of the first 10 such genes. We can see that some might
show some meaningful changes between 60 and 72H, but others clearly just have
a single replicate that is different or increased variability at 72H that results in a lack of significance in 72H.

```{r showMissingDEGenes, echo=FALSE, fig.cap="The expression values over time for 10 genes found significant DE in timepoints 48,60,120,168H, but not in the 72H timepoint."}
exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
plot_splines_data(data, moanin_model, subset_data=exampleGenes,
		  colors=ann_colors$Group, smooth=TRUE)
```

@@

`moanin` extends this idea by not only fitting a splines function for each gene, but also providing functionality to compare time course
data between different treatment conditions, using a similar mechanism of
contrasts -- only now the contrasts are differences between the estimated mean
functions. This is done with the function `DE_timecourse`, which takes as
similar input that of `DE_timepoints`, only now will fit spline functions for each gene and test the entire mean
function (and unlike `DE_timecourse` therefore does not require the extra step of expanding the contrasts into contrasts for individual timepoints).

@@

# The function takes the data (data.frame or named matrix), the meta data
# (data.frame containing a timepoint and group column, the first corresponding
# to the time-course information, the latter corresponding to the
# treatment).
DE_results = DE_timecourse(
    data, moanin_model, timecourse_contrasts,
    use_voom_weights=FALSE)
```

The output from `DE_timecourse` is a matrix of (raw) p-values and Benjamini-Hochberg corrected q-values for each comparison.

```{r}
names(DE_results)
```

For convenience we will separate these into two matrices.

```{r}
pval_columns = colnames(DE_results)[
    grepl("pval", colnames(DE_results))]
qval_columns = colnames(DE_results)[
    grepl("qval", colnames(DE_results))]
pvalues = DE_results[, pval_columns]
qvalues = DE_results[, qval_columns]
```

@@

```{r echo=FALSE, fig.cap="Distribution of DE genes per condition based on global splines analysis"}
num_de = colSums(qvalues < 0.05)
par(mar=c(7.1,5.1,1.1,1.1))
names(num_de) = c(
    "Kawasaki", "California", "Vietnam LD", "Vietname HD")
barplot(num_de, col="black", ylab="Number of DE genes",las=2)
```

### Log-fold Change for Time Course Data

The next step in a classical differential expression analysis is typically to
assess the effect of the treatment by calculating for each gene the log fold change in the gene expression between the treatment and control.

Computing the log fold change on a time-course experiment is not trivial: one
can be interested in the average log-fold change across time, or the
cumulative log-fold change. Sometimes a gene can be over-expressed at the
beginning of the time-course data, and then under-expressed at the end of the
experiment. As a result, `moanin` provides a number of possible ways to
compute the log fold change across the whole time-course. This is done via the function `estimate_log_fold_change` which takes as
arguments the data, the `moanin_model` object, the contrasts to evaluate, and the method to use to estimate the log-fold change.

**Individual Timepoints** The first method ("timely") gives a simple interface to compute the log-fold change
for each individual timepoints.

@@

**Cumulative Effect** Sometimes, a single value per gene for each contrast is more useful, and `estimate_log_fold_change` used above provides several options for this as well. Here is a
table of the possible ways `estimate_log_fold_change` can compute log-fold change values (including `timely` discussed above).

| Name   | Formula   |     |
|------- | --------- |--- |
| timely | $\text{lfc}(t)$  | Per time point |
| sum  | $\sum_t \text{lfc}(t)$  | Sum of log fold change. |
| abs_sum  | $\sum_t \lvert \text{lfc}(t)\lvert$  | Always positive |
| max  | $\max_t \lvert \text{lfc}(t) \lvert$  | Always positive  |
| min  | $\min_t \lvert \text{lfc}(t) \lvert$  | Always positive |
| timecourse | See details below | Captures overall strength *and* direction  |

The method "timecourse" tries to capture the overall strength and direction of the response in the following way: we leverage the
timepoint by timepoint log-fold change $\text{lfc}(t)$, and apply the
following formula:

@@

Note, however, that when a gene is not consistently up or down-regulated the estimation
of the direction will not represent accurately the changes observed.

We demonstrate several of these methods on our data.

@@

```{r display_lfc, echo=FALSE}
kable(log_fold_change_timecourse[1:5, ], digits=3, booktabs=TRUE)
```

In Figure \ref{fig:compareLogFoldChange} we plot the log-fold change summary of each of these methods against each other. We see that each methods captures different elements of
the time-course data, for example, overall change versus the largest change.

```{r compareLogFoldChange, echo=FALSE, fig.height=6, fig.width=6, message=FALSE, fig.cap="Comparison of the different methods of calculating log-fold change per gene"}
par(mfrow=c(2, 2))
plot(log_fold_change_timecourse[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="timecourse vs sum",
     xlab="timecourse", ylab="sum")
plot(log_fold_change_timecourse[, "M-C"],
     log_fold_change_max[, "M-C"], pch=20, main="timecourse vs max",
     xlab="timecourse", ylab="max")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_min[, "M-C"], pch=20, main="max vs min",
     xlab="max", ylab="min")
plot(log_fold_change_max[, "M-C"],
     log_fold_change_sum[, "M-C"], pch=20, main="max vs sum",
     xlab="max", ylab="sum")
```

With a single measure of log-fold change and the p-value, we can now look
at the traditional volcano plot. In Figure \ref{fig:volcano}, we show the example of a volcano plot for the comparaison
of the control to the Kawasaki strain, using the "timecourse" method of calculating log fold change.

```{r volcano, fig.height=4, fig.width=4, fig.cap="Volcano plot for the comparison of the Kawasaki strain to control, where fold-change is calculated with the timecourse method"}
pvalue = DE_results[, "M-K_pval"]
names(pvalue) = row.names(DE_results)
lfc_timecourse = log_fold_change_timecourse[, "M-K"]
names(lfc_timecourse) = row.names(log_fold_change_timecourse)

@@

### Visualizing Genes of Interest

The package `moanin` also provides a simple utility function (`plot_splines_data`) to
visualize gene time-course data from different conditions. In Figure \ref{fig:topPvalueGenes}, we plot the 10
genes with the smallest p-values.

```{r topPvalueGenes, fig.height=6, fig.width=6}
top_DE_genes_pval = names(sort(pvalue)[1:10])
plot_splines_data(data[top_DE_genes_pval, ], moanin_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))
```

For each gene, the individual data points are plotted against time and color coded by their condition. Further, a fitted spline function for each group is plotted to aid in comparing global trends across conditions.

Figure \ref{fig:topLogFold}, we visualize the genes with the largest absolute timecourse log-fold
change.

```{r topLogFold,fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_timecourse),
	 decreasing=TRUE)[1:10])
plot_splines_data(data[top_DE_genes_lfc, ], moanin_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))

@@

In examining these visualizations, we can see that genes often follow similar
patterns of expression, although on a different scale for each gene. We can
leverage this observation to cluster the genes into groups of similar patterns
of transcriptomic response.

@@

The very large number of genes found differentially expressed impair any
interpretation one would attempt: with 70\% of the genome found differentially
expressed, all pathways are affected by the treatment. Hence the next step of
the workflow to cluster gene expression according to their dynamical response
to the treatment.

**Filtering** Before clustering the genes, we first reduce the set of
genes of interest to genes that are (1) found significantly differentially
expressed; (2) a large-fold change between conditions. Reducing the set of genes on which to perform the clustering allows to
estimate the centroids of the clusters with more stability.

To do this, we first
aggregate all p-values obtained during the time-course differential expresison
step in a single p-value using Fisher's method [@fisher:statistical] (`pvalues_fisher_method`).
Next we
select all the genes which have a Fisher-adjusted p-value below 0.05 and a log-fold 
change of at least two for at least one condition and one time-point.

@@

### Clustering Based on Spline Fits

After filtering, we are left with `r dim(y)[1]` genes. We can then apply a
clustering routine. As observed by looking at genes found differentially expressed, 
many genes share a similar gene expression pattern, but on different scale.

We thus propose the following adaptation of k-means:

- **Splines estimation**: for each gene, fit the splines function with the basis
  of your choice (as contained in the `moanin_model` object).
- **Rescaling splines**: for each gene, rescale the estimated splines function
  such that the values are bounded between 0 and 1 and thus comparable between genes.
- **K-means**: apply k-means on the rescaled fitted values of the splines to estimate the
  cluster centroids.

The rescaling of the splines is similar to the rescaling genes by their standard deviation, which is common for gene expression studies without a time component.

These clustering steps are performed by the `splines_kmeans`
function in `moanin`. For now, we will set the number of clusters to be $8$, though we
will return to the question of picking the best number of clusters below.

@@

- `centroids`: a matrix containing the cluster centroids. The matrix is of
  shape (`n_centroid`, `n_samples`).
- `clusters`: a vector of size `n_genes`, containing the cluster assignments given by the `kmeans` step of each gene

We then use the `plot_splines_data`  function, only now applied to the centroids, to visualize the centroids of
each cluster obtained with the splines k-means model.

@@

These centroids are on a 0-1 scale, because of our rescaling of the spline
fits, and do not represent the actual gene expression level. In Figure \ref{fig:genesVsCentroids}, we plot a few
of the actual genes assigned to cluster 2 to see the difference:

```{r genesVsCentroids, fig.height=6, fig.width=6}
cluster_of_interest = 2
cluster2Genes = names(
    kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
dataToPlot=rbind(y[cluster2Genes[3:6], ],"Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
plot_splines_data(dataToPlot, moanin_model,
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

The spline plots are fitted per gene in the above plot and illustrate that
the genes assigned to a cluster have differing degrees to which their data
fits that of the cluster centroid, and indeed fits that of their individual
spline fits.

@@

Because there is variability in how well the genes fit a cluster, we would like to be able to score how well a gene fits a cluster. Furthermore, we arbitrarily chose a subset of genes based on our filter, and we would
like to have a mechanism to assign all genes to a cluster.

Thus, the next step in the clustering portion of the workflow is a scoring and label step. Each gene is given a score that
corresponds to a goodness-of-fit between each gene and each cluster, computed
as follows:

@@

The score then allows us to assign all genes to a cluster (i.e. a label) based on the cluster for which they have the best score, regardless of whether the gene was used in the clustering procedure.

The scoring and labeling is done via the `splines_kmeans_score_and_label`
function. This function calculates the goodness of fit of the gene to the
cluster centroid and gives a cluster label to the gene if they have a sufficiently high score, as we explain above.

@@

    - `scores`: the matrix of shape `n_cluster` x `n_genes`,
      containing for each gene and each cluster, the goodness of fit as
      described above.
    - `labels`: the labels for all of the genes with a
      sufficiently good goodness-of-fit score.

**Assigning cluster labels:** We could just assign each gene to a cluster based on which cluster gave the minimum score. By default, `splines_kmeans_score_and_label` does not do that, but rather requires a sufficiently low enough goodness-of-fit score. The criteria for being "sufficiently low" is based on looking at the distribution of the scores of all genes on all clusters (i.e. the entire `scores` matrix returned by `splines_kmeans_score_and_label`). A gene is then assigned to a cluster only if their best score is above the 50% percentile of that distribution, with the remaining genes getting `NA` as their assignment. Note that because the distribution of scores of genes across *all* clusters are used, this is not equivalent to assigning 50% of genes to a cluster -- it is possible that all genes are assigned to a cluster.

@@

After running `scores_and_labels` on our data, we now have `r length(labels)` genes that
are assigned to a cluster.

We can visualize the impact of this filtering process by considering the distribution of goodness-of-fit scores for each
cluster if we did not have the filtering cutoff of `splines_kmeans_score_and_label`, i.e. if we simply assign every gene to a cluster based on which gives them their minimum score.
We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure.

```{r visualizeScores, fig.height=6, fix.width=6, echo=TRUE, fig.cap="Distribution of goodness-of-fit scores for each cluster. The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled by \\texttt{splines_kmeans_score_and_label}."}
# Get the best score and best label for all of the genes
# This is different from output of `splines_kmeans_score_and_label`, which filters
scores = rowMin(scores_and_labels$scores)
labels = apply(scores_and_labels$scores, 1, which.min)
max_score = quantile(scores, 0.5)
par(mfrow=c(3, 3))
n_clusters = dim(kmeans_clusters$centroids)[1]
for(cluster_id in 1:n_clusters){
    hist(scores[labels == cluster_id],
	 breaks=(1:50/50), xlim=c(0, 1),
	 col="black", main=paste("C", cluster_id, sep=""),
	 xlab="score", ylab="Num. genes")
    abline(v=max_score, col="red", lwd=3, lty=2)
}

Note that for some of the genes, their scores in all clusters is 1. Such genes fit
poorly to all clusters and the assignment we did above to one cluster is done arbitrarily to the first matching cluster.
This underlines the importance of filtering genes that fit poorly to clusters.

**EAP** In fact, `which.min` which you used above, picks the first match, so actually a preference to assign into first cluster in this plot.

We can also investigate the
differences between the labels provided by the splines k-means model and our
scoring and labeling step, in terms of the number of genes assigned to each cluster (Figure \ref{fig:kmeansVsScore}.

```{r kmeansVsScore, echo=FALSE, fig.height=6, fig.width=6, warning=FALSE, fig.cap="Number of genes assigned to each cluster based on kmeans criteria of nearest centroid (top) versus our goodness-of-fit filtering strategy (bottom)"}
par(mfrow=c(2, 1))
barplot(table(kmeans_clusters$clusters),
	col="black",
	main="labeling from splines k-means")
barplot(table(labels), col="black", main="labeling from scores_and_labels function")
```
